#Team 6: Paula, Sadiksha**
  
# 🧬 Capstone Project: R-based Mass Spectrometry Proteomics Workflow
  
#This document provides a step-by-step guide for your capstone project on analyzing proteomics data using R.

---
  
## 🔁 Workflow Overview
  ### 1. 📁 Dataset Acquisition  
  #**Input:** PRIDE PXD Identifier  
 
#(*Example: `PXD0123456`, accessible via [PRIDE](https://www.ebi.ac.uk/pride/)*)

#Use the `rpx` package to access metadata and download a dataset that includes `.mzID` files from a published study.
```{r}
library(rpx)
px <- PXDataset("PXD054726")  #male
pxfiles(px) ## Retrieve and display a list of all files associated with the PXDataset object 'px'

```

```{r}
files <- pxfiles(px) #Retrieve and display a list of all files associated with the PXDataset object 'px'
mzid_files <- pxget(px, grep("mzid", pxfiles(px))) #Identify the .mzID files from the retrieved files
```


---
  
### 2. 🧱 PSM Object Creation & Preprocessing  
#**Goal:** Generate a PSM (Peptide-Spectrum Match) object from `.mzID` files  
 
#- Convert `.mzID` files into `PSM` objects
# Assess:
# Number of decoy hits
# Score distributions
# PSM rank
# Apply filtering based on FDR or identification score

```{r}
library(PSMatch)
library(ggplot2)
```


```{r}
psms <- PSM(mzid_files, parser = 'mzID') #Convert .mzID files into PSM objects
```

```{r}
names(psms)
```
```{r}
# PSM rank
#    * Number of decoy hits
#no isdecoy column
library("dplyr")
psm_df <- tidyr::as_tibble(psms)
psm_df
#Converted into dataframe because it is just easier to check in environment and easier to work on for me. 
sum(is.na(psm_df$description)) #0 #one way is to check if there are proteins without description. -> No NAs in description
#Another idea we discussed was that if the peak score is low (for eg 0), they could be removed as decoy. So I will come back to that in later step after score distribution
```


```{r}
#    * Number of decoy hits
#no isdecoy column
#Converted into tibble because it is just easier to check in environment and easier to work on for me. as  hist script :) 
sum(is.na(psm_df$description)) #0 #one way is to check if there are proteins without description. -> No NAs in description
#Another idea we discussed was that if the peak score is low (for eg 0), they could be removed as decoy. So I will come back to that in later step after score distribution
```



```{r}
#Score distributions
# Score distributions
score_distribution <- summary(psms@listData$`mascot:score`)
print(score_distribution)
#hist(psms@listData$`mascot:score`, breaks = 100) #one way to do histogram

table(psm_df$passthreshold) #true#76701  may be after removing decoys or delta score or mascot score thereshold? 

library(ggplot2)
ggplot(psm_df, aes(x = mascot.score)) +
  geom_density()+
   labs(title = "Density Plot of Mascot Scores",
       x = "Mascot Score",
       y = "Density") +
  theme_minimal() 
```
```{r}

#when filtering one plan is to remove everything at 0. just exploring where to filter
ggplot(psm_df, aes(x = mascot.score)) +
  geom_density(fill = "blue", alpha = 0.5) +  # Fill color for the density plot
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +  # Vertical line at x = 1
  labs(title = "Density Plot of Mascot Scores",
       x = "Mascot Score",
       y = "Density") +
  theme_minimal() 
```



```{r}
# PSM rank
psm_rank_distribution <- table(psm_df$rank)
print(psm_rank_distribution) 
```

```{r}
#Apply filtering based on FDR or identification score ?
psms <- psms %>% filterPsmRank() #keeping only those with rank 1 #skip this?
psms <- filterPsmFdr(psms, FDR = 0.05, fdr = psmVariables(psms)["fdr"], verbose = TRUE) #if we had fdr column, we don't have that column
psm_df <- tidyr::as_tibble(psms)
orig_df <- nrow(psm_df)
psm_df <- psm_df %>% filter(mascot.score > 0) #filtering based on identification score becuase we do not have decoys adn fdr in this data
filtered_df <- nrow(psm_df)
number_of_decoys <- orig_df -filtered_df
(number_of_decoys) 
```

```{r}
adjacencyMatrix(psms)
```





  
### 3. 🧬 Protein & Peptide Identification  
# **Goal:** Determine identified peptides and proteins  
#Count the number of identified peptides and proteins
#Review peptide-to-protein mapping:
#Razor proteins
#Protein groups
```{r}
describePeptides(psms) #number of peptides in unfiltered dataset

dim(psm_df)
nrow(psm_df)
length(unique(psm_df$pepseq)) #peptides #76701  after filtering
length(unique(psm_df$accession)) #proteins? #598 number of proteins
```

```{r}
adj <- makeAdjacencyMatrix(psms) #this one is without the mascot.score <0 removed because i dont know how to convert tibble to psm again. but rank 1 is filtered. 
dim(adj)
```
```{r}
length(unique(psms$sequence))
length(unique(psms$DatabaseAccess))
cc <- ConnectedComponents(adj)
length(cc)
cc
(i <- which(nrows(cc) > 2 & ncols(cc) > 2))
dims(cc)[i, ]
cx <- connectedComponents(cc, 372) #plotting this for simplicity because this component has low number of rows 17 rows(fragments?)
cx
plotAdjacencyMatrix(cx)
plotAdjacencyMatrix(cx, 1)

```
```{r}
cc
```

```{r}
#Among the 2617 and the 662 proteins, we have 542 connected components. 

#226 correspond to single proteins identified by single peptide. 
#0 represent protein groups identified by a single shared peptide
#260 represent single proteins identified by multiple unique peptides:
#Finally, arguable those that warrant additional exploration are those that are composed of multiple peptides and multiple proteins. There are 56 thereof in this identification

#which means there are 662 unique sequences.
#there are no unique database accesses recorded in the dataset.
#there are 542 connected components in the graph.
#The line (i <- which(nrows(cc) > 2 & ncols(cc) > 2)) identifies the indices of connected components that have more than 2 rows and more than 2 columns.
#The output dims(cc)[i, ] shows the dimensions (number of rows and columns) of these selected connected components
#12 components have more than 2 rows and more than 2 columns, with varying sizes.
```


```{r}
#trying to find razor proteins
```

---
  
### 4. 🔄 QFeature Aggregation (Optional)  
# **Goal:** Use quantification data if available  
  
#Use `aggregateFeatures()` from the `QFeatures` package to aggregate:
  #PSMs ➡️ Peptides ➡️ Proteins

library(QFeatures)
library(dplyr)  
qfeatures <- QFeatures(list(PSM = SummarizedExperiment(assays = list(lognorm_peptides = as.matrix(id)))))
# Assuming 'peptide' is the column with peptide sequences and 'Mascot:score' is the scoring metric

rowData(qfeatures[["PSM"]])
qfeatures[["PSM"]]
data("b1370p17_S9_Tot_Prot_P31_FRAX_24h_1000nM_3.RAW.-1.mgf")

qfeatures <- aggregateFeatures(qfeatures,
                               i = "PSM",  # Input assay
                               name = "Peptides",  # Name of the new aggregated feature
                               fcol = "peptide",  # Column containing peptide sequences
                               fun = colMeans,  # Aggregation function (mean of scores, for example)
                               na.rm = TRUE)

library(MSnbase)



---
  
  ### 5. 🧼 Normalization & Imputation (Optional)  
  **Goal:** Correct for technical variation and handle missing values  
↓  
- Normalize using `normalize()` or `normalize_vsn()`
- Impute missing values with `impute()`

---
  
  ### 6. 🧪 Protein Inference & Quantification (Optional)  
  **Goal:** Summarize and quantify proteins  
↓  
- Aggregate peptide-level intensities into protein-level quantities
- Optionally annotate protein IDs using external databases (e.g., UniProt)

---
  
  ### 7. 📊 Statistical Analysis (Optional)  
  **Goal:** Identify differentially abundant proteins  
↓  
- Perform statistical testing (e.g., using `test_diff()`)
- Filter by:
  - Log2 fold-change
- Adjusted p-value (e.g., FDR < 0.05)

---
  
  ### 8. 📈 Visualization & Export (Optional)  
  **Goal:** Visualize data for interpretation  
↓  
- Create visual summaries:
  - PCA plots
- Volcano plots
- Heatmaps  
- Use packages such as `ggplot2`, `DEP`, or `limma`

---
  
  ### 9. 📖 Comparison with Published Results  
  **Goal:** Benchmark your findings  
↓  
Create a comparison table including:
  - Number of spectra identified
- Number of peptides and proteins
- Key figures or results (if available)
- Comments on reproducibility

---
  
  ### 10. 📝 Final Report & Interpretation  
  **Goal:** Reflect on your analysis  
↓  
Submit the following:
  - Source code (R script or RMarkdown) in your team folder
- A 1–2 page report that includes:
  - Background of the dataset and study
- Summary table of key results
- Discussion:
  > Why do your results match or differ from the original publication?
  
  ---
  
  ✅ **Tip:** Convert your `.Rmd` into a clean `README.md` using `knitr::knit("README.Rmd")`.
